\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-18 paper/manuscript.qmd
mention metad in intro

Decision-making is often conceptualized as involving distinct processes: the choice itself, and the confidence associated with that choice [@sanders2016signatures]. 
In experimental paradigms, these can be somewhat disentangled by prompting participants to make a discrete choice selection (e.g., "True" *vs.* "False"), followed by a separate confidence rating. 
However, this artificial separation makes its joint analysis difficult, and may not reflect real-world scenarios, where individuals often express both choice and confidence simultaneously using a single, continuous scale. In such scales, each side can represent a distinct latent category, and the distance from the midpoint can indicate the level of confidence or certainty.
This integrated response format typically results in bimodal distributions, with peaks corresponding to the mean confidence on either side.
Traditional beta regression models, which assume unimodal distributions within the (0,1) interval, are ill-suited for such data. One alternative is to transform the data into two variables a posteriori: binarizing the side to represent choice and calculating the absolute distance from the midpoint to represent confidence. 
These can then be modeled separately, for instance, using logistic regressions for choice and beta regressions for confidence [see @makowski2025too for an example]. 
While this approach can provide additional insights into underlying mechanisms compared to a unique model, it assumes psychological and statistical independence between choice and confidence, which may not hold true in practice.

@@

To model data of subjective scales in which the left and right sides can be conceptualized as two different choices (e.g., True/False, Agree/Disagree, etc.) and the magnitude of the response (how much the cursor is set away from the mid-point) as the confidence, we introduce the Choice-Confidence (CHOCO) model (@fig-two).
It consists of a three‐part mixture on $x \in [0, 1]$:

>>>>

Decision-making is often conceptualized as involving distinct processes: the choice itself, and the confidence associated with that choice [@sanders2016signatures]. 
In experimental paradigms, these can be somewhat disentangled by prompting participants to make a discrete choice selection (e.g., correct *vs.* error), followed by a separate confidence rating, which has led to the development of bespoke approaches to model the underlying data-generation process [e.g., @fleming2017hmeta; @rausch2021modelling].
However, this artificial separation may not reflect real-world scenarios, where individuals express simultaneously through a single scale both a discrete choice and an analog judgment of confidence, certainty, or degree of agreement (note that we will use the term "confidence" hereafter as a general term, without assuming a specific cognitive process). In instances where each side represents a distinct latent category (e.g., True/False, Agree/Disagree, Positive/Negative) - as opposed to a truly unimodal dimension, the absolute distance from the midpoint could be conceptualized as a indicating a continuous degree of "confidence".
This integrated response format typically results in bimodal distributions, with peaks corresponding to the mean confidence on either side.
Traditional beta regression models, which assume unimodal distributions within the (0,1) interval, are ill-suited for such data. One alternative is to transform the data into two variables a posteriori: binarizing the side to represent choice and calculating the absolute distance from the midpoint to represent confidence. 
These can then be modeled separately, for instance, using logistic regressions for choice and beta regressions for confidence [see @makowski2025too for an example]. 
While this approach can provide additional insights into underlying mechanisms compared to a unique model, it assumes psychological and statistical independence between choice and confidence, which may not hold true in practice.

@@

To model data of subjective scales in which the left and right sides can be conceptualized as two different choices and the magnitude of the response (how much the cursor is set away from the mid-point) as the confidence, we introduce the Choice-Confidence (CHOCO) model (@fig-two).
It consists of a three‐part mixture on $x \in [0, 1]$:

### 2025-06-18 paper/manuscript.tex
mention metad in intro

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, opacityback=0, colframe=quarto-callout-note-color-frame, arc=.35mm, colback=white, left=2mm, rightrule=.15mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm]

@@

Decision-making is often conceptualized as involving distinct processes:
the choice itself, and the confidence associated with that choice
(\citeproc{ref-sanders2016signatures}{Sanders et al., 2016}). In
experimental paradigms, these can be somewhat disentangled by prompting
participants to make a discrete choice selection (e.g., ``True''
\emph{vs.} ``False''), followed by a separate confidence rating.
However, this artificial separation makes its joint analysis difficult,
and may not reflect real-world scenarios, where individuals often
express both choice and confidence simultaneously using a single,
continuous scale. In such scales, each side can represent a distinct
latent category, and the distance from the midpoint can indicate the
level of confidence or certainty. This integrated response format
typically results in bimodal distributions, with peaks corresponding to
the mean confidence on either side. Traditional beta regression models,
which assume unimodal distributions within the (0,1) interval, are
ill-suited for such data. One alternative is to transform the data into
two variables a posteriori: binarizing the side to represent choice and
calculating the absolute distance from the midpoint to represent
confidence. These can then be modeled separately, for instance, using
logistic regressions for choice and beta regressions for confidence (see
\citeproc{ref-makowski2025too}{Makowski, Te, et al., 2025} for an
example). While this approach can provide additional insights into
underlying mechanisms compared to a unique model, it assumes
psychological and statistical independence between choice and
confidence, which may not hold true in practice.

@@

To model data of subjective scales in which the left and right sides can
be conceptualized as two different choices (e.g., True/False,
Agree/Disagree, etc.) and the magnitude of the response (how much the
cursor is set away from the mid-point) as the confidence, we introduce
the Choice-Confidence (CHOCO) model (Figure~\ref{fig-two}). It consists
of a three‐part mixture on \(x \in [0, 1]\):

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, arc=.35mm, colback=white, rightrule=.15mm, leftrule=.75mm, left=2mm, breakable, colframe=quarto-callout-note-color-frame, opacityback=0, toprule=.15mm]

@@

Decision-making is often conceptualized as involving distinct processes:
the choice itself, and the confidence associated with that choice
(\citeproc{ref-sanders2016signatures}{Sanders et al., 2016}). In
experimental paradigms, these can be somewhat disentangled by prompting
participants to make a discrete choice selection (e.g., correct
\emph{vs.} error), followed by a separate confidence rating, which has
led to the development of bespoke approaches to model the underlying
data-generation process (e.g., \citeproc{ref-fleming2017hmeta}{Fleming,
2017}; \citeproc{ref-rausch2021modelling}{Rausch et al., 2021}).
However, this artificial separation may not reflect real-world
scenarios, where individuals express simultaneously through a single
scale both a discrete choice and an analog judgment of confidence,
certainty, or degree of agreement (note that we will use the term
``confidence'' hereafter as a general term, without assuming a specific
cognitive process). In instances where each side represents a distinct
latent category (e.g., True/False, Agree/Disagree, Positive/Negative) -
as opposed to a truly unimodal dimension, the absolute distance from the
midpoint could be conceptualized as a indicating a continuous degree of
``confidence''. This integrated response format typically results in
bimodal distributions, with peaks corresponding to the mean confidence
on either side. Traditional beta regression models, which assume
unimodal distributions within the (0,1) interval, are ill-suited for
such data. One alternative is to transform the data into two variables a
posteriori: binarizing the side to represent choice and calculating the
absolute distance from the midpoint to represent confidence. These can
then be modeled separately, for instance, using logistic regressions for
choice and beta regressions for confidence (see
\citeproc{ref-makowski2025too}{Makowski, Te, et al., 2025} for an
example). While this approach can provide additional insights into
underlying mechanisms compared to a unique model, it assumes
psychological and statistical independence between choice and
confidence, which may not hold true in practice.

@@

To model data of subjective scales in which the left and right sides can
be conceptualized as two different choices and the magnitude of the
response (how much the cursor is set away from the mid-point) as the
confidence, we introduce the Choice-Confidence (CHOCO) model
(Figure~\ref{fig-two}). It consists of a three‐part mixture on
\(x \in [0, 1]\):
